Rhizaria in the oligotrophic ocean exhibit clear temporal and vertical variability.

### 2024-06-26 manuscript/main.tex
render manuscript with revisions

Rhizaria are an extremely diverse super-group of single-celled organisms
which occupy a wide range of habitats. Planktonic Rhizaria include
Foraminifera, Radiolaria (including Acantharea and Collodaria) and
Phaeodaria, all of which are observed throughout the global ocean (Biard
et al., 2016; Laget et al., 2024) . While the taxonomy of these
organisms has recently undergone several reclassifications (Biard,
2022a), their presence in ocean ecosystems has been long known to
oceanographers. Some of the earliest records of their existence are from
oceanographic expeditions in the 19th century (Haekel, 1887). Rhizaria
are unique members of the plankton and protist community because they
can reach large sizes (up to several mm in diameter) and they construct
intricate mineral skeletons out of either silica, strontium, or calcium
carbonate (Biard, 2022a; Kimoto, 2015; Nakamura and Suzuki, 2015; Suzuki
and Not, 2015). Despite their noticeable morphology and global
distribution, Rhizaria were largely understudied throughout the 20th
century. Fragile organisms like Rhizaria are difficult to adequately
study because they can be destroyed through standard zooplankton
sampling techniques and preserve poorly. A number of studies in the late
1900s did employ alternative techniques to quantify Rhizaria including
diaphragm pumps (Michaels, 1988) or blue-water SCUBA collections (Bijma
et al., 1990; Caron et al., 1995; Caron and Be, 1984). However, the bulk
of Rhizaria research was constrained to sediment traps or
paleontological studies of sediment (Boltovskoy et al., 1993; Takahashi
et al., 1983).

Recently, the advent of molecular techniques and \emph{in situ} imaging
tools have ignited a renewed focus on Rhizaria in pelagic ecosystems
(Caron, 2016). Molecular tools have shed light on Rhizaria taxonomy
(Biard, 2022a), distribution (Blanco-Bercial, 2020; Sogawa et al.,
2022), ecology (Decelle et al., 2012; Nakamura et al., 2023),
biogeochemical roles (Gutierrez-Rodriguez et al., 2019; Laget et al.,
2024), and metabolism (Cohen et al., 2023). Yet, despite the excellent
taxonomic resolution provided by molecular approaches, they do not
provide a truly quantitative metric for estimating Rhizaria abundance or
biomass. \emph{In situ} imaging tools however, offer the ability to
observe organisms in the natural state and quantify their abundance
(Barth and Stone, 2024; Ohman, 2019). Biard et al. (2016) utilized
\emph{in situ} imaging at a global scale to suggest large
(\textgreater500\(\mu\)m diameter) Rhizaria were substantial
contributors to the ocean carbon standing stock. While more recent
calculations suggest lower carbon contribution (Laget et al., 2024),
Rhizaria nonetheless have substantial influences on biogeochemical
cycling. This influence is due to their unique mineral skeletons and
ability to repackage small particles into larger, fast sinking ones
(Ikenoue et al., 2019). Thus, some Rhizaria may play substantial roles
in ocean export through their interception of sinking particles (Laget
et al., 2024; Stukel et al., 2018). However, Rhizaria ecology is poorly
understood (Biard, 2022a) and their ecological roles are likely more
than diverse than simple particle interception.

The ecological role of Rhizaria in plankton communities is complicated
due to the fact different taxa can exhibit very different trophic modes.
Many Rhizaria are strictly heterotrophic (Biard, 2022a), yet their
feeding modes can be quite varied. Phaeodaria are largely thought to be
flux-feeders, collecting and feeding on sinking particles (Gowing,
1989). Alternatively, Retaria can be either exclusively heterotrophic or
mixotrophic, utilizing photosynthetic algal symbionts (Anderson, 2014;
Decelle et al., 2015). Mixotrophic Foraminifera host a variety of
endosymbiont partners (Decelle et al., 2015; Lee, 2006), which are
thought to support early and adult life stages and significantly
contribute to total primary productivity (Kimoto, 2015). Still,
Foraminifera are omnivorous, possibly even predominately carnivorous,
with several studies suggesting that they can be effective predators
(Anderson and Bé, 1976; Gaskell et al., 2019), mainly consuming live
copepods (Caron and Be, 1984). Radiolaria have several lineages, many of
which host symbionts (Biard, 2022b). Amongst Radiolaria, arguably the
most widespread are Collodaria who can be either large solitary cells or
form massive colonies, up to several meters in length (Swanberg and
Anderson, 1981). All known Collodaria species host dinoflagellate
symbionts (Biard, 2022b) and can contribute substantially to primary
productivity, particularly in oligotrophic ocean regions (Caron et al.,
1995; Dennett et al., 2002). This Collodaria-symbiont association has
been suggested as a reason for their high abundances throughout the
photic zone of oligotrophic environments globally (Biard et al., 2017,
2016). A few Acantharea (Radiolaria order) clades host algal symbionts
(Biard, 2022b; Decelle et al., 2012), notably with two clades forming an
exclusive relationship with \emph{Phaeocystis}. However, globally,
Acantharea are less abundant than Collodaria (Biard, 2022a) and
contribute less to total primary productivity (Michaels et al., 1995).
This may be due to the fact several clades of Acantharea are
cyst-forming and strictly heterotrophs (Biard, 2022b; Decelle et al.,
2013). Furthermore, Mars Brisbin et al. (2020) documented apparent
predation behavior in Acantharea near the surface, suggesting that there
may be a larger reliance on carnivory.

Given the high abundances, yet diverse trophic strategies found among
Rhizaria taxa, it is reasonable to expect some form of niche
partitioning. A number of studies do suggest evidence for vertical
zonation between Rhizaria groups according to various trophic
strategies. Taxon-specific studies of Radiolaria suggest they may be
restricted to the euphotic zone (Boltovskoy, 2017; Michaels, 1988).
Although some studies report Acantharea in deeper waters (Decelle et
al., 2013; Gutiérrez-Rodríguez et al., 2022), Phaeodaria, alternatively,
are generally found beyond depths where light can sustain
photosynthesis, but they can feed on sinking particles (Laget et al.,
2024; Stukel et al., 2018). In an imaging-based study of the whole
Rhizaria community, Biard and Ohman (2020) noted clear patterns in
vertical zonation which largely corresponded to different trophic roles.
Protists communities, including Rhizaria, have been described to
partition along autotroph-mixotroph-heterotroph gradients with
increasing depth in the water column (Blanco-Bercial et al., 2022; Laget
et al., 2024). Yet, few studies have made direct attempts to relate
Rhizaria abundances to environmental factors (Biard and Ohman, 2020). In
part, this is due to the fact few studies have been able to sample
Rhizaria in the same location over a consistent timeframe (Boltovskoy et
al., 1993; Gutiérrez-Rodríguez et al., 2022; Hull et al., 2011; Michaels
et al., 1995; Michaels, 1988). Furthermore, very few studies have
utilized \emph{in situ} imaging, arguably the best method for
quantifying Rhizaria, to consistently throughout the full mesopelagic
(Laget et al., 2024) . Given this lack of information, there are many
unknowns with respect to Rhizaria ecology, seasonality and phenology
across different groups.

@@

Data were collected in collaboration with the Bermuda Atlantic
Time-series Study (Lomas et al., 2013; Michaels and Knap, 1996) on board
the R/V Atlantic Explorer. Cruises were conducted at approximately
monthly intervals. Rhizaria individuals were sampled using the
Underwater Vision Profiler 5 (UVP) (Picheral et al., 2010). The UVP5 is
an in-situ camera used to capture plankton and particle images and has
been well established to accurately quantify large Rhizaria abundance
(Barth and Stone, 2022; Biard et al., 2016; Biard and Ohman, 2020; Drago
et al., 2022; Llopis Monferrer et al., 2022; Panaïotis et al., 2023;
Stukel et al., 2019; Stukel et al., 2018). The UVP5 was mounted to the
sampling rosette and collected data autonomously on routine casts, from
which only the downcast data are utilized. The UVP5 was deployed from
June-September 2019 then from October 2020 - January 2022, during which
time the BATS region was sampled for 3-5 days at monthly intervals (see
sampling details in Supplemental Table 1). Casts were filtered to only
include data collected in the BATS region, far offshore of Bermuda in
the Sargasso Sea (approximately 31.0\(^oN\)-32.5\(^oN\),
64.25\(^oW\)-63\(^oW\); Supplemental Figure 1). In general, casts
extended to either 200m, 500m, or 1200m deep, with a few extended into
the bathypelagic (4500m). However, Rhizaria were only typically found in
large abundances throughout the epipelagic and mesopelagic zones. As
such, we limit this study to results from the upper 1000m of the water
column.

A variety of biotic and abiotic data were collected during each BATS
cruise. The UVP5 provided particle count data at a high-frequency from
each cast. Particle concentration, as a proxy for prey field (Whitmore
and Ohman, 2021), was calculated from this data for all particles
184\(\mu m\) - 900\(\mu m\). The lower size range was set by what could
be reliably sampled by the UVP5's pixel resolution (\textgreater2px;
0.092mm per pixel) and the upper size range was set to below the size of
observable Rhizaria yet large enough to be inclusive of particles on
which some Rhizaria may feed. For each UVP cast supporting continuous
profiles of the CTD parameters salinity, temperature, and auxiliary CTD
channels; Dissolved Oxygen (DO), \emph{in situ} chlorophyll fluorescence
were measured at 24Hz using the BATS CTD package. On select casts,
Niskin bottles were used to collect bacterial abundance estimates (via
epifluorescence microscopy) as well as measure inorganic nutrients
(\(NO_3\), and \(Si\) as silicate/sicilic acid) at discrete depths. On
each cruise, flux estimates of total mass, organic carbon, and nitrogen
were also collected using sediment traps; in the present study we
utilized flux to the mesopelagic as the flux at 200m. Also primary
productivity was estimated through measuring \(^{14}C\) uptake rates
from \emph{in situ} incubations. For full descriptions of the BATS
sampling program and methods, see Knap et al. (1997) and Lomas et al.
(2013) for a review. Additionally, data can be viewed online
\href{https://bats.bios.asu.edu/bats-data/}{(https://bats.bios.asu.edu/bats-data/)}.

@@

Individual vignettes of Rhizaria images were identified using the
classification platform EcoTaxa (Picheral et al., n.d.). Data were
pre-sorted utilizing a random-forest classifier and pre-trained learning
set. Taxonomic classification were done based on morphological
parameters measured in Zooprocess (Gorsky et al., 2010), which includes
parameters such as major axis, equivalent spherical diameter (ESD), and
grey values. While there are sparse taxonomic guides for \emph{in situ}
images of Rhizaria, identification largely relied on descriptions in
(Nakamura and Suzuki, 2015; Suzuki and Not, 2015; and Biard and Ohman,
2020). Using the aforementioned sources and publicly available EcoTaxa
projects, we constructed a guide accessible at:
\url{https://thealexbarth.github.io/media//Project_Items/Oligotrophic_Community/ecotaxa_UVP-guide-stone-lab.pdf}.
Broadly, Rhizaria were classified as Foraminifera, Radiolaria
(Acantharea or Collodaria), or as a variety of Phaeodaria families
(Figure 1). When identification could not be confidently made between a
few candidate taxa, a less specific label was used. As a result, we have
data from ``unidentified Rhizaria'', which typically were vignettes not
distinguishable between Aulacanthidae or Acantharea or ``unidentified
Phaeodaria'', which are clearly Phaeodaria but not distinguishable into
a family.

@@

The UVP5 samples at \textasciitilde15Hz rate as it descends the water
column and records the exact position of each particle larger than
600\(\mu m\) (ESD). However, identified rhizaria ranged from a
934\(\mu m\) ESD Aulacanthidae cell to a Collodarian colony over 10mm in
diameter. To confirm that the UVP5 was sampling adequately across all
size ranges, an normalized biomass size spectrum (NBSS) slope was
constructed to identify a drop-off which would indicate poor-sampling at
the small size range (Barth and Stone, 2024; Lombard et al., 2019).
However, it was evident from this analysis that all size ranges were
adequately sampled across the size range (Supplemental Figure 2) so no
data were excluded. The UVP5 reports the exact depth at which a particle
is recorded, however to estimate abundance, observations must be binned
over fixed depth intervals. Our deployments had variable descent depths
and speeds with more casts descending to 500m than 1000m and descents
quicker through the epipelagic than the mesopelagic (see Barth and Stone
(2022) for an extended discussion of UVP5 data processing). For the
present study, Rhizaria abundances were estimated in 25m vertical bins,
which offer a moderate sampling volume per bin (average 0.948\(m^3\) in
the epipelagic and 0.589\(m^3\) in the mesopelagic; Supplemental Table
1) while still maintaining ecologically relevant widths. However,
concentrations in a 25m bin would need to be greater than 2.428 ind.
\(m^{-3}\) and 3.912 ind. \(m^{-3}\), in the epipelagic and mesopelagic
respectively, to fall below a 10\% non-detection risk (Barth and Stone,
2024; Benfield et al., 1996). Because we typically observed many
Rhizaria taxa below these concentrations, we present the 25m binned data
to visualize broad-scale average distributions. For quantifying and
modelling Rhizaria abundances, we present integrated abundance
estimates, with each cast. Due to the variable descent depths of the
UVP, data are categorized as epipelagic (0-200m), upper mesopelagic
(200-500m), and lower mesopelagic (500-1000m). The average sampling
volume integrated through these regions were 7.59\(m^2\), 7.06\(m^2\),
and 11.77\(m^2\), with non-detection thresholds at 0.30 ind. \(m^{-2}\),
0.33 ind. \(m^{-2}\), and 0.20 ind. \(m^{-2}\) respectively. All UVP
data processing was done using the \texttt{EcotaxaTools} package in R
(Barth 2023).

@@

While the environmental conditions were generally low in variation and
oligotrophic, there is considerable influence from deep winter-mixing
and summer stratification as well as secondary influences from mesoscale
eddies which result in spatiotemporal heterogeniety (Lomas et al., 2013;
McGillicuddy et al., 1998). Variability in the water column structure
was visible during the study period (Figure 2). This is best evidenced
through the temperature profiles; In the late summer and early fall
there was a stratified water column with high temperatures in the
surface (\textless75m) (Figure 2A) and slightly elevated salinity
(Figure 2B). This warm, stratified period appeared more intense during
the few months sampled in 2019. In 2021, we observed the stratified
layer slowly dissipated into the winter months following mixed layer
entrainment. There was a consistent oxygen minimum zone (OMZ) located at
about 800m deep (Figure 2C). February 2021 saw a notable downwelling
event, likely due to a passing anti-cyclonic eddy which impacted the
local region during early 2021. During this phase, warmer, oxic water
was significantly depressed deeper into the mesopelagic. This process
was reversed in the spring months (March, April) primary due to the
interaction of convective mixing and the passing of a strong cyclonic
eddy resulting in a deep cold mixed layer. Primary production was
highest during the spring mixing period, evidenced both by \emph{in
situ} fluorescence (Figure 2D) and productivity incubation experiments
(Figure 3A). Originating near the surface, the productivity peak moved
deeper throughout the spring and declined into the summer (Figure 2D).
However, there was a notable, yet smaller productivity bump in the late
summer and early fall (Figure 3A) which occurred deeper in the
epipelagic (Figure 2D). The particle concentration (184\(\mu m\) -
900\(\mu m\)) was closely coupled to chlorophyll-a patterns.

@@

{\centering \includegraphics{images/02-environmental.pdf}

@@

@@

In the epipelagic Rhizaria exhibited a notable seasonal pattern.
Rhizaria abundances were higher in the summer months and lower during
the winter. During a prior time period, Blanco-Bercial et al. (2022)
noted that there is considerable seasonality in the community
composition of all protists. Despite the seasonality of Rhizaria
abundance, community composition was relatively consistent, with
Collodaria representing the bulk of the community. It should be noted
that the overall taxonomic resolution of the UVP5 is fairly low, so
there may be a switching of species within the broad groups identified
in this study which were not captured. Throughout the mesopelagic,
month-to-month variation in 2021 was relatively low. Again, this is
consistent with observations from metabarcoding of the whole protist
community in the same study region (Blanco-Bercial et al., 2022). This
finding is not surprising as the overall seasonal variation in
environmental conditions in this region were low.

Overall Rhizaria were the most commonly identified group of mesoplankton
throughout the study period. We do note that the UVP5 commonly captures
\emph{Trichodesmium} colonies, yet these were excluded in this
comparison as they are strictly autotrophs. It should be noted that
previous work has suggested that avoidance behavior with the UVP is
possible, at times likely, for visual and highly mobile zooplankton
(Barth and Stone, 2022). Thus, the percent contribution reported here
(42.7\%) of Rhizaria to the total mesozooplankton community may be
inflated due to under sampling of organisms such as Euphausiids and
Chaetognaths which have quick escape responses. Regardless, it is worth
noting that in the same region, with data collected in 2012 and 2013
using similar calculation methods, Biard et al. (2016) estimated large
Rhizaria only contribute 15\% of the total mesozooplankton community in
the upper 500m. Likely, Rhizaria display considerable interannual
variability. In the present study, we noticed considerably higher
Rhizaria abundance throughout the water column in 2019 compared to 2021.
Similarly, in the California Current a multi-year study of
Aulosphaeridae saw considerable variation between years (Biard et al.,
2018). To truly understand the drivers of interannual variability
however, sustained observations of Rhizaria over longer time periods are
required.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-anderson2014}
Anderson, O.R., 2014. Living Together in the Plankton: A Survey of
Marine Protist Symbioses. Acta Protozoologica 53, 29--38.
\url{https://doi.org/10.4467/16890027AP.13.0019.1116}

>>>>

Rhizaria are an extremely diverse super-group of single-celled organisms
which occupy a wide range of habitats. Planktonic Rhizaria include
Foraminifera, Radiolaria (including Acantharea and Collodaria) and
Phaeodaria, all of which are observed throughout the global ocean (Biard
\emph{et al.}, 2016; Laget \emph{et al.}, 2024) . While the taxonomy of
these organisms has recently undergone several reclassifications (Biard,
2022b), their presence in ocean ecosystems has been long known to
oceanographers. Some of the earliest records of their existence are from
oceanographic expeditions in the 19th century (Haekel, 1887). Rhizaria
are unique members of the plankton and protist community because they
can reach large sizes (up to several mm in diameter) and they construct
intricate mineral skeletons out of either silica, strontium, or calcium
carbonate (Kimoto, 2015; Nakamura and Suzuki, 2015; Suzuki and Not,
2015; Biard, 2022b). Despite their noticeable morphology and global
distribution, Rhizaria were largely understudied throughout the 20th
century. Fragile organisms like Rhizaria are difficult to adequately
study because they can be destroyed through standard zooplankton
sampling techniques and preserve poorly. A number of studies in the late
1900s did employ alternative techniques to quantify Rhizaria including
diaphragm pumps (Michaels, 1988) or blue-water SCUBA collections (Caron
and Be, 1984; Bijma \emph{et al.}, 1990; Caron \emph{et al.}, 1995).
However, the bulk of Rhizaria research was constrained to sediment traps
or paleontological studies of sediment (Takahashi \emph{et al.}, 1983;
Boltovskoy \emph{et al.}, 1993).

Recently, the advent of molecular techniques and \emph{in situ} imaging
tools have ignited a renewed focus on Rhizaria in pelagic ecosystems
(Caron, 2016). Molecular tools have shed light on Rhizaria taxonomy
(Biard, 2022b), distribution (Blanco-Bercial, 2020; Sogawa \emph{et
al.}, 2022), ecology (Decelle \emph{et al.}, 2012; Nakamura \emph{et
al.}, 2023), biogeochemical roles (Gutierrez-Rodriguez \emph{et al.},
2019; Laget \emph{et al.}, 2024), and metabolism (Cohen \emph{et al.},
2023). Yet, despite the excellent taxonomic resolution provided by
molecular approaches, they do not provide a truly quantitative metric
for estimating Rhizaria abundance or biomass. \emph{In situ} imaging
tools however, offer the ability to observe organisms in the natural
state and quantify their abundance (Ohman, 2019; Barth and Stone, 2024).
Biard \emph{et al.} (2016) utilized \emph{in situ} imaging at a global
scale to suggest large (\textgreater500\(\mu\)m diameter) Rhizaria were
substantial contributors to the ocean carbon standing stock. While more
recent calculations suggest lower carbon contribution (Laget \emph{et
al.}, 2024), Rhizaria nonetheless have substantial influences on
biogeochemical cycling. This influence is due to their unique mineral
skeletons and ability to repackage small particles into larger, fast
sinking ones (Ikenoue \emph{et al.}, 2019). Thus, some Rhizaria may play
substantial roles in ocean export through their interception of sinking
particles (Stukel \emph{et al.}, 2018; Laget \emph{et al.}, 2024).
However, Rhizaria ecology is poorly understood (Biard, 2022b) and their
ecological roles are likely more than diverse than simple particle
interception.

The ecological role of Rhizaria in plankton communities is complicated
due to the fact different taxa can exhibit very different trophic modes.
Many Rhizaria are strictly heterotrophic (Biard, 2022b), yet their
feeding modes can be quite varied. Phaeodaria are largely thought to be
flux-feeders, collecting and feeding on sinking particles (Gowing,
1989). Alternatively, Retaria can be either exclusively heterotrophic or
mixotrophic, utilizing photosynthetic algal symbionts (Anderson, 2014;
Decelle \emph{et al.}, 2015). Mixotrophic Foraminifera host a variety of
endosymbiont partners (Lee, 2006; Decelle \emph{et al.}, 2015), which
are thought to support early and adult life stages and significantly
contribute to total primary productivity (Kimoto, 2015). Still,
Foraminifera are omnivorous, possibly even predominately carnivorous,
with several studies suggesting that they can be effective predators
(Anderson and Bé, 1976; Gaskell \emph{et al.}, 2019), mainly consuming
live copepods (Caron and Be, 1984). Radiolaria have several lineages,
many of which host symbionts (Biard, 2022a). Amongst Radiolaria,
arguably the most widespread are Collodaria who can be either large
solitary cells or form massive colonies, up to several meters in length
(Swanberg and Anderson, 1981). All known Collodaria species host
dinoflagellate symbionts (Biard, 2022a) and can contribute substantially
to primary productivity, particularly in oligotrophic ocean regions
(Caron \emph{et al.}, 1995; Dennett \emph{et al.}, 2002). This
Collodaria-symbiont association has been suggested as a reason for their
high abundances throughout the photic zone of oligotrophic environments
globally (Biard \emph{et al.}, 2016, 2017). A few Acantharea (Radiolaria
order) clades host algal symbionts (Decelle \emph{et al.}, 2012; Biard,
2022a), notably with two clades forming an exclusive relationship with
\emph{Phaeocystis}. However, globally, Acantharea are less abundant than
Collodaria (Biard, 2022b) and contribute less to total primary
productivity (Michaels \emph{et al.}, 1995). This may be due to the fact
several clades of Acantharea are cyst-forming and strictly heterotrophs
(Decelle \emph{et al.}, 2013; Biard, 2022a). Furthermore, Mars Brisbin
\emph{et al.} (2020) documented apparent predation behavior in
Acantharea near the surface, suggesting that there may be a larger
reliance on carnivory.

Given the high abundances, yet diverse trophic strategies found among
Rhizaria taxa, it is reasonable to expect some form of niche
partitioning. A number of studies do suggest evidence for vertical
zonation between Rhizaria groups according to various trophic
strategies. Taxon-specific studies of Radiolaria suggest they may be
restricted to the euphotic zone (Michaels, 1988; Boltovskoy, 2017).
Although some studies report Acantharea in deeper waters (Decelle
\emph{et al.}, 2013; Gutiérrez-Rodríguez \emph{et al.}, 2022),
Phaeodaria, alternatively, are generally found beyond depths where light
can sustain photosynthesis, but they can feed on sinking particles
(Stukel \emph{et al.}, 2018; Laget \emph{et al.}, 2024). In an
imaging-based study of the whole Rhizaria community, Biard and Ohman
(2020) noted clear patterns in vertical zonation which largely
corresponded to different trophic roles. Protists communities, including
Rhizaria, have been described to partition along
autotroph-mixotroph-heterotroph gradients with increasing depth in the
water column (Blanco-Bercial \emph{et al.}, 2022; Laget \emph{et al.},
2024). Yet, few studies have made direct attempts to relate Rhizaria
abundances to environmental factors (Biard and Ohman, 2020). In part,
this is due to the fact few studies have been able to sample Rhizaria in
the same location over a consistent timeframe (Michaels, 1988;
Boltovskoy \emph{et al.}, 1993; Michaels \emph{et al.}, 1995; Hull
\emph{et al.}, 2011; Gutiérrez-Rodríguez \emph{et al.}, 2022).
Furthermore, very few studies have utilized \emph{in situ} imaging,
arguably the best method for quantifying Rhizaria, to consistently
throughout the full mesopelagic (Laget \emph{et al.}, 2024) . Given this
lack of information, there are many unknowns with respect to Rhizaria
ecology, seasonality and phenology across different groups.

@@

Data were collected in collaboration with the Bermuda Atlantic
Time-series Study (Michaels and Knap, 1996; Lomas \emph{et al.}, 2013)
on board the R/V Atlantic Explorer. Cruises were conducted at
approximately monthly intervals. Rhizaria individuals were sampled using
the Underwater Vision Profiler 5 (UVP) (Picheral \emph{et al.}, 2010).
The UVP5 is an in-situ camera used to capture plankton and particle
images and has been well established to accurately quantify large
Rhizaria abundance (Biard \emph{et al.}, 2016; Stukel \emph{et al.},
2018; Stukel \emph{et al.}, 2019; Biard and Ohman, 2020; Barth and
Stone, 2022; Drago \emph{et al.}, 2022; Llopis Monferrer \emph{et al.},
2022; Panaïotis \emph{et al.}, 2023). The UVP5 was mounted to the
sampling rosette and collected data autonomously on routine casts, from
which only the downcast data are utilized. The UVP5 was deployed from
June-September 2019 then from October 2020 - January 2022, during which
time the BATS region was sampled for 3-5 days at monthly intervals (see
sampling details in Supplemental Table 1). Casts were filtered to only
include data collected in the BATS region, far offshore of Bermuda in
the Sargasso Sea (approximately 31.0\(^oN\)-32.5\(^oN\),
64.25\(^oW\)-63\(^oW\); Supplemental Figure 1). In general, casts
extended to either 200m, 500m, or 1200m deep, with a few extended into
the bathypelagic (4500m). However, Rhizaria were only typically found in
large abundances throughout the epipelagic and mesopelagic zones. As
such, we limit this study to results from the upper 1000m of the water
column.

A variety of biotic and abiotic data were collected during each BATS
cruise. The UVP5 provided particle count data at a high-frequency from
each cast. Particle concentration, as a proxy for prey field (Whitmore
and Ohman, 2021), was calculated from this data for all particles
184\(\mu m\) - 900\(\mu m\). The lower size range was set by what could
be reliably sampled by the UVP5's pixel resolution (\textgreater2px;
0.092mm per pixel) and the upper size range was set to below the size of
observable Rhizaria yet large enough to be inclusive of particles on
which some Rhizaria may feed. For each UVP cast supporting continuous
profiles of the CTD parameters salinity, temperature, and auxiliary CTD
channels; Dissolved Oxygen (DO), \emph{in situ} chlorophyll fluorescence
were measured at 24Hz using the BATS CTD package. On select casts,
Niskin bottles were used to collect bacterial abundance estimates (via
epifluorescence microscopy) as well as measure inorganic nutrients
(\(NO_3\), and \(Si\) as silicate/sicilic acid) at discrete depths. On
each cruise, flux estimates of total mass, organic carbon, and nitrogen
were also collected using sediment traps; in the present study we
utilized flux to the mesopelagic as the flux at 200m. Also primary
productivity was estimated through measuring \(^{14}C\) uptake rates
from \emph{in situ} incubations. For full descriptions of the BATS
sampling program and methods, see Knap \emph{et al.} (1997) and Lomas
\emph{et al.} (2013) for a review. Additionally, data can be viewed
online
\href{https://bats.bios.asu.edu/bats-data/}{(https://bats.bios.asu.edu/bats-data/)}.

@@

Individual vignettes of Rhizaria images were identified using the
classification platform EcoTaxa (Picheral \emph{et al.}). Data were
pre-sorted utilizing a random-forest classifier and pre-trained learning
set. Taxonomic classification were done based on morphological
parameters measured in Zooprocess (Gorsky \emph{et al.}, 2010), which
includes parameters such as major axis, equivalent spherical diameter
(ESD), and grey values. While there are sparse taxonomic guides for
\emph{in situ} images of Rhizaria, identification largely relied on
descriptions in (Nakamura and Suzuki, 2015; Suzuki and Not, 2015; and
Biard and Ohman, 2020). Using the aforementioned sources and publicly
available EcoTaxa projects, we constructed a guide accessible at:
\url{https://thealexbarth.github.io/media//Project_Items/Oligotrophic_Community/ecotaxa_UVP-guide-stone-lab.pdf}.
Broadly, Rhizaria were classified as Foraminifera, Radiolaria
(Acantharea or Collodaria), or as a variety of Phaeodaria families
(Figure 1). When identification could not be confidently made between a
few candidate taxa, a less specific label was used. As a result, we have
data from ``unidentified Rhizaria'', which typically were vignettes not
distinguishable between Aulacanthidae or Acantharea or ``unidentified
Phaeodaria'', which are clearly Phaeodaria but not distinguishable into
a family.

@@

The UVP5 samples at \textasciitilde15Hz rate as it descends the water
column and records the exact position of each particle larger than
600\(\mu m\) (ESD). However, identified rhizaria ranged from a
934\(\mu m\) ESD Aulacanthidae cell to a Collodarian colony over 10mm in
diameter. To confirm that the UVP5 was sampling adequately across all
size ranges, an normalized biomass size spectrum (NBSS) slope was
constructed to identify a drop-off which would indicate poor-sampling at
the small size range (Lombard \emph{et al.}, 2019; Barth and Stone,
2024). However, it was evident from this analysis that all size ranges
were adequately sampled across the size range (Supplemental Figure 2) so
no data were excluded. The UVP5 reports the exact depth at which a
particle is recorded, however to estimate abundance, observations must
be binned over fixed depth intervals. Our deployments had variable
descent depths and speeds with more casts descending to 500m than 1000m
and descents quicker through the epipelagic than the mesopelagic (see
Barth and Stone (2022) for an extended discussion of UVP5 data
processing). For the present study, Rhizaria abundances were estimated
in 25m vertical bins, which offer a moderate sampling volume per bin
(average 0.948\(m^3\) in the epipelagic and 0.589\(m^3\) in the
mesopelagic; Supplemental Table 1) while still maintaining ecologically
relevant widths. However, concentrations in a 25m bin would need to be
greater than 2.428 ind. \(m^{-3}\) and 3.912 ind. \(m^{-3}\), in the
epipelagic and mesopelagic respectively, to fall below a 10\%
non-detection risk (Benfield \emph{et al.}, 1996; Barth and Stone,
2024). Because we typically observed many Rhizaria taxa below these
concentrations, we present the 25m binned data to visualize broad-scale
average distributions. For quantifying and modelling Rhizaria
abundances, we present integrated abundance estimates, with each cast.
Due to the variable descent depths of the UVP, data are categorized as
epipelagic (0-200m), upper mesopelagic (200-500m), and lower mesopelagic
(500-1000m). The average sampling volume integrated through these
regions were 7.59\(m^2\), 7.06\(m^2\), and 11.77\(m^2\), with
non-detection thresholds at 0.30 ind. \(m^{-2}\), 0.33 ind. \(m^{-2}\),
and 0.20 ind. \(m^{-2}\) respectively. All UVP data processing was done
using the \texttt{EcotaxaTools} package in R (Barth 2023).

@@

While the environmental conditions were generally low in variation and
oligotrophic, there is considerable influence from deep winter-mixing
and summer stratification as well as secondary influences from mesoscale
eddies which result in spatiotemporal heterogeniety (McGillicuddy
\emph{et al.}, 1998; Lomas \emph{et al.}, 2013). Variability in the
water column structure was visible during the study period (Figure 2).
This is best evidenced through the temperature profiles; In the late
summer and early fall there was a stratified water column with high
temperatures in the surface (\textless75m) (Figure 2A) and slightly
elevated salinity (Figure 2B). This warm, stratified period appeared
more intense during the few months sampled in 2019. In 2021, we observed
the stratified layer slowly dissipated into the winter months following
mixed layer entrainment. There was a consistent oxygen minimum zone
(OMZ) located at about 800m deep (Figure 2C). February 2021 saw a
notable downwelling event, likely due to a passing anti-cyclonic eddy
which impacted the local region during early 2021. During this phase,
warmer, oxic water was significantly depressed deeper into the
mesopelagic. This process was reversed in the spring months (March,
April) primary due to the interaction of convective mixing and the
passing of a strong cyclonic eddy resulting in a deep cold mixed layer.
Primary production was highest during the spring mixing period,
evidenced both by \emph{in situ} fluorescence (Figure 2D) and
productivity incubation experiments (Figure 3A). Originating near the
surface, the productivity peak moved deeper throughout the spring and
declined into the summer (Figure 2D). However, there was a notable, yet
smaller productivity bump in the late summer and early fall (Figure 3A)
which occurred deeper in the epipelagic (Figure 2D). The particle
concentration (184\(\mu m\) - 900\(\mu m\)) was closely coupled to
chlorophyll-a patterns.

@@

{\centering \includegraphics{images/02_environmental.pdf}

@@

{\centering \includegraphics{images/06_tot-partials.pdf}

\caption{Partial effect plots from Generalized Additive Models (GAMs) of
all Rhizaria abundance. Models were constructed separately for depth
integrated abundance through the Epipelagic (0-200m) (A-G), Upper
Mesopelagic (200-500m) (H-J), and the Lower Mesopelagic (500-1000m)
(K-L). Each panel represents the partial effect of a GAM model term,
with smoothed relationship shown shaded with 95\% confidence intervals.
Individual points are depth integrated values for all Rhizaria from an
individual cast, colored by month.}

@@

In the epipelagic Rhizaria exhibited a notable seasonal pattern.
Rhizaria abundances were higher in the summer months and lower during
the winter. During a prior time period, Blanco-Bercial \emph{et al.}
(2022) noted that there is considerable seasonality in the community
composition of all protists. Despite the seasonality of Rhizaria
abundance, community composition was relatively consistent, with
Collodaria representing the bulk of the community. It should be noted
that the overall taxonomic resolution of the UVP5 is fairly low, so
there may be a switching of species within the broad groups identified
in this study which were not captured. Throughout the mesopelagic,
month-to-month variation in 2021 was relatively low. Again, this is
consistent with observations from metabarcoding of the whole protist
community in the same study region (Blanco-Bercial \emph{et al.}, 2022).
This finding is not surprising as the overall seasonal variation in
environmental conditions in this region were low.

Overall Rhizaria were the most commonly identified group of mesoplankton
throughout the study period. We do note that the UVP5 commonly captures
\emph{Trichodesmium} colonies, yet these were excluded in this
comparison as they are strictly autotrophs. It should be noted that
previous work has suggested that avoidance behavior with the UVP is
possible, at times likely, for visual and highly mobile zooplankton
(Barth and Stone, 2022). Thus, the percent contribution reported here
(42.7\%) of Rhizaria to the total mesozooplankton community may be
inflated due to under sampling of organisms such as Euphausiids and
Chaetognaths which have quick escape responses. Regardless, it is worth
noting that in the same region, with data collected in 2012 and 2013
using similar calculation methods, Biard \emph{et al.} (2016) estimated
large Rhizaria only contribute 15\% of the total mesozooplankton
community in the upper 500m. Likely, Rhizaria display considerable
interannual variability. In the present study, we noticed considerably
higher Rhizaria abundance throughout the water column in 2019 compared
to 2021. Similarly, in the California Current a multi-year study of
Aulosphaeridae saw considerable variation between years (Biard \emph{et
al.}, 2018). To truly understand the drivers of interannual variability
however, sustained observations of Rhizaria over longer time periods are
required.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-anderson2014}
Anderson, O. R. (2014)
\href{https://doi.org/10.4467/16890027AP.13.0019.1116}{Living Together
in the Plankton: A Survey of Marine Protist Symbioses}. \emph{Acta
Protozoologica}, \textbf{53}, 29--38.
